---
title: "Regulon analysis"
author: "Yi Xie"
date: "2023-06-08"
output:
  pdf_document: default
  html_document:
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

The following analysis is replicating what described in 'Regulon analaysis' in Material and Method session (https://www.biorxiv.org/content/10.1101/2022.12.20.521178v1?rss=1)

Path to original figure pdf files:
https://drive.google.com/drive/u/1/folders/1iX1HGcKyitPYyQjCG501GW5xJmd5F1EW

## 1. INSTALL pySCENIC
```{r, engine = 'bash', eval = FALSE}
module load singularity
cd /data/petretto/home/e0205142/software/
singularity build aertslab-pyscenic-0.12.0.sif docker://aertslab/pyscenic:0.12.0
```

Usage of pySCENIC in singulatiry container:
```{r, engine = 'bash', eval = FALSE}
## Overall usage
singularity run aertslab-pyscenic-0.12.0.sif pyscenic -h
## sub-command help
singularity run aertslab-pyscenic-0.12.0.sif pyscenic grn -h
singularity run aertslab-pyscenic-0.12.0.sif pyscenic ctx -h
```

## 2. Dependency
Download dependent databases of SCENIC including:
- TF lists
- human genome ranked by regulatory features (feather files)
- Motif-to-TF annotations

```{r, engine = 'bash', eval = FALSE}
mkdir pyscenic_dep
cd /pyscenic_dep

# A. list of human TFs:
https://github.com/aertslab/pySCENIC/blob/master/resources/hs_hgnc_tfs.txt

# B. database ranking the whole human genome on regulatory features (TFs etc) (.feather)
## B.1 500bp upstream of TSS (transcripton starting site)
feather_database_url='https://resources.aertslab.org/cistarget/databases/
homo_sapiens/hg19/refseq_r45/mc9nr/gene_based/
hg19-500bp-upstream-10species.mc9nr.genes_vs_motifs.rankings.feather'
wget "${feather_database_url}" --no-check-certificate
### B.2 5kb around the TSS (10kb in total)
feather_database_url='https://resources.aertslab.org/cistarget/databases/
homo_sapiens/hg19/refseq_r45/mc9nr/gene_based/
hg19-tss-centered-5kb-10species.mc9nr.genes_vs_motifs.rankings.feather'
wget "${feather_database_url}" --no-check-certificate

## C. Motif to TF annotations:
wget --no-check-certificate
https://resources.aertslab.org/cistarget/motif2tf/motifs-v9-nr.hgnc-m0.001-o0.0.tbl

```

## 3. Loom file
Input to pySCENIC is loom file. Extract count martix from SeuratObject and convert it with python module loompy

Enter my working directory for this analysis
```{r, engine = 'bash', eval = FALSE}
cd /data/petretto/home/e0205142/project/Kevin/fig4
```
### 3.1 Extract raw count matrix from SeuratObject
expression matrix from integrated assay (24105 cells by 3000 genes)
```{r, eval = FALSE}
SPP1seu <- readRDS('./data/SPP1seu.rds')
write.csv(t(as.matrix(SPP1seu@assays$integrated@data)),
          file = "./data/SPP1seu_exp.csv")
cellInfo <- SPP1seu@meta.data
head(cellInfo)
write.csv(cellInfo, file = "./data/SPP1seu_cellInfo.csv")
```

### 3.2 Convert to loom file
```{r, engine = 'bash', eval = FALSE}
cd ./code/
vim make_loom.py

## add following python codes to make_loom.py
import os, sys
os.getcwd()
os.listdir(os.getcwd())
import loompy as lp
import numpy as np
import scanpy as sc

x=sc.read_csv("../data/SPP1seu_exp.csv");
row_attrs = {"Gene": np.array(x.var_names),};
col_attrs = {"CellID": np.array(x.obs_names)};
lp.create("../data/SPP1seu.loom",x.X.transpose(),row_attrs,col_attrs)

## excute make_loom.py
conda activate torch-py38 # I have loompy and scanpy pre-installed in this conda env
python make_loom.py
conda deactivate
```

## 4. run pySCENIC

Start an interactive session in HPC for following analysis
```{r, engine = 'bash', eval = FALSE}
qsub -IX -l select=1:ncpus=50:mem=200G:ngpus=0 -l walltime=72:00:00 -q super

cd /data/petretto/home/e0205142/project/Kevin/fig4/
module load singularity/

```

### 4.1 grnboost2
Infer GRN with grnboost2. This step took around 2.5 hrs to run.
```{r, engine = 'bash', eval = FALSE}
singularity run ~/software/aertslab-pyscenic-0.12.0.sif \
pyscenic grn --num_workers 40 \
  --sparse \
  --method grnboost2 \
  --output ./final_output/scenic/SPP1seu.adj.csv \
  ./data/SPP1seu.loom \
  ~/software/pyscenic_dep/hs_hgnc_tfs.txt

```

### 4.2 RcisTarget
Start to filter GRNs to get regulons with RcisTarget.

Here two databases (up 500bp & ±5kb) are used for motif enrichment as described in MAM manuscript.


```{r, engine = 'bash', eval = FALSE}

singularity run /data/petretto/home/e0205142/software/aertslab-pyscenic-0.12.0.sif \
pyscenic ctx --num_workers 40 \
  --mode "dask_multiprocessing" \
  --all_modules \
  --min_genes 10 \
  --output ./final_output/scenic/SPP1seu.regulons.csv \
  --expression_mtx_fname ./data/SPP1seu.loom \
  --annotations_fname /data/petretto/home/e0205142/software/pyscenic_dep/motifs-v9-nr.hgnc-m0.001-o0.0.tbl \
  ./final_output/scenic/SPP1seu.adj.csv \
  /data/petretto/home/e0205142/software/pyscenic_dep/hg19-500bp-upstream-10species.mc9nr.genes_vs_motifs.rankings.feather \
  /data/petretto/home/e0205142/software/pyscenic_dep/hg19-tss-centered-5kb-10species.mc9nr.genes_vs_motifs.rankings.feather

```

### 4.3 AUCell
To score each regulon's activity in each cell and get regulon activity matrix.
```{r, engine = 'bash', eval = FALSE}
singularity run /data/petretto/home/e0205142/software/aertslab-pyscenic-0.12.0.sif \
pyscenic aucell --num_workers 40 \
  --output ./final_output/scenic/SPP1seu_SCENIC_res.loom \
  ./data/SPP1seu.loom \
  ./final_output/scenic/SPP1seu.regulons.csv

```

